Would NEMO benefit from using the \psyclone\ system?

### 2015-11-20 docs/NEMO/psyclone_and_nemo.tex
NEMO workplan: added content to WP on infrastructure and ref. for parco2015 paper

\psyclone\ has been designed from the ground up with performance and
performance portability in mind. Rather than automatically optimising
code, as compilers do, \psyclone\ allows an optimisation expert to
specify a set of optimisations in a script, which it then applies to
the code. The \psykal separation of concerns can be considered a
stripping back of many of the optimisations that have been hard coded
into a program and \psyclone\ then allows such optimisations to be
specified in a script via a set of transformations (a recipe of
transformations if you like) to apply to the code. This hopefully
means that there is no (or at least minimal) loss in performance
compared to hand optimised code when using \psyclone.

@@

NEMO currently uses an $(i,j,k)$ array-index ordering and on current
(CPU-based) architectures reaches a strong-scaling limit when the MPI
sub-domains are about $40\times 40$ grid points. In contrast, the
number of vertical levels in current configurations is typically
$\sim75$. In addition it is likely that the majority of loops over
levels in NEMO are parallelisable. It is therefore highly desirable to
be able to exploit this and this lends weight to option (2). In order
for the PSy layer to manage the iteration space in k it must be
possible to formulate a complete set of rules for the different types
of k loops in the whole of NEMO. For instance, it must be possible to
infer the value of the lower bound of the k-loop around a given kernel
from a knowledge of the grid-point type of the field being updated by
a kernel (plus other meta-data describing the kernel). The number of
exceptions to this must be small if an effective Domain Specific
Language is to be developed for \psyclone\'s application to NEMO.

@@

\section{Index re-reordering}

\psyclone\ does not directly support the idea of re-ordering NEMO's data
layout. This question has been asked of \psyclone\ before as it is
thought that i,j,k may work best on some architectures and k,i,j on
others. However, the separation of science code and optimisations and
parallelism combined with the separation of the i,j,k looping
structure from the Kernel science code should make it much more
tractable to manually evaluate the benefits of data reordering. In
future work we would be interested in exploring the possibility of
supporting data re-ordering within \psyclone\ in cases where the PSy
layer controls the full iteration space.

\section{Suggested workplan - 1 Week}

@@

\section{Suggested workplan - 1 year}

@@

\subsection{WP3: Infrastructure support}
\label{wp3_infrastructure}

Field objects or arrays?

The existing NEMO code base is already MPI parallel and therefore any
introduction of \psyclone\ must work within this framework.

@@

Performance is a key issue for NEMO and is one of the reasons for
considering \psyclone\ -- the resulting code will no longer be
restricted to one particular architecture. However, a key first step
is to ensure that the introduction of the \psykal\ separation of
concerns does not harm performance on current machines. Or, if it
does, then that performance can be recovered through code
transformations. We have shown that this is the case for two
shallow-water models~\cite{XXX} but this work must be extended for the
3D parts of NEMO. Benchmarking of the configurations chosen in WP1
will inform the key routines used to examine this issue. Note that
this benchmarking must be performed on processor counts typical of
those used in production runs.

@@

>>>>

\psyclone\ has been designed from the ground up with performance and
performance portability in mind. Rather than automatically optimising
code, as compilers do, \psyclone\ allows an optimisation expert to
specify a set of optimisations in a script, which it then applies to
the code. The \psykal\ separation of concerns can be considered a
stripping back of many of the optimisations that have been hard coded
into a program and \psyclone\ then allows such optimisations to be
specified in a script via a set of transformations (a recipe of
transformations if you like) to apply to the code. This hopefully
means that there is no (or at least minimal) loss in performance
compared to hand optimised code when using \psyclone.

@@

NEMO currently uses an $(i,j,k)$ array-index ordering and on current
(CPU-based) architectures reaches a strong-scaling limit when the MPI
sub-domains are about $40\times 40$ grid points. In contrast, the
number of vertical levels in current configurations is typically
$\sim75$. In addition it is likely that the majority of loops over
levels in NEMO are parallelisable. It is therefore highly desirable to
be able to exploit this and this lends weight to option (2). In order
for the PSy layer to manage the iteration space in k it must be
possible to formulate a complete set of rules for the different types
of k loops in the whole of NEMO. For instance, it must be possible to
infer the value of the lower bound of the k-loop around a given kernel
from a knowledge of the grid-point type of the field being updated by
a kernel (plus other meta-data describing the kernel). The number of
exceptions to this must be small if an effective Domain Specific
Language is to be developed for \psyclone's application to NEMO.

@@

\section{Index Re-ordering}

\psyclone\ does not directly support the idea of re-ordering NEMO's data
layout. This question has been asked of \psyclone\ before as it is
thought that $i,j,k$ may work best on some architectures and $k,i,j$ on
others. However, the separation of science code and optimisations and
parallelism combined with the separation of the $i,j,k$ looping
structure from the Kernel science code should make it much more
tractable to manually evaluate the benefits of data reordering. In
future work we would be interested in exploring the possibility of
supporting data re-ordering within \psyclone\ in cases where the PSy
layer controls the full iteration space.

\section{Suggested Workplan - 1 Week}

@@

\section{Suggested Workplan - 1 Year}

@@

\subsection{WP3: Infrastructure Support}
\label{wp3_infrastructure}

\psyclone\ must interface to some sort of infrastructure that provides
support for manipulating fields - e.g. creating them and performing
halo swaps. Obviously NEMO already contains support for performing
domain decomposition and MPI operations. Any incremental introduction
of \psyclone\ must therefore work within this framework.

In the latest versions of GOcean and Dynamo the infrastructure is
built around the concept of field objects. While these have benefits,
such an approach would be a significant change to NEMO and may also
conflict with the use of AGRIF. In the initial stages of this work it
will therefore probably be necessary for the infrastructure to work
with fields as Fortran arrays rather than derived types.

@@

Performance is a key issue for NEMO and is one of the reasons for
considering \psyclone\ -- the resulting code will no longer be
restricted to one particular architecture. However, a key first step
is to ensure that the introduction of the \psykal\ separation of
concerns does not harm performance on current machines. Or, if it
does, then that performance can be recovered through code
transformations. We have shown that this is the case for two
shallow-water models~\cite{psykal_shallow} but this work must be
extended for the 3D parts of NEMO. Benchmarking of the configurations
chosen in WP1 will inform the key routines used to examine this
issue. Note that this benchmarking must be performed on processor
counts typical of those used in production runs.

@@

\begin{thebibliography}{1}

\bibitem{psykal_shallow} A.~R.~Porter, R.~W.~Ford, M.~Ashworth, G.~D.~Riley and M.~Modani, ``Towards Compiler-Agnostic Performance in Finite-Difference Codes'', to appear in the Proceedings of ParCo 2015.

\end{thebibliography}
